Constraining the Interiors of Asteroids Through Close Encounters

### 2022-04-06 paper/main.tex
Draft zero

\jtd{Write}
The last numbered section should briefly summarise what has been done, and describe
the final conclusions which the authors draw from their work.

@@

The physical parameters of the asteroid are kept the same as the Earth encounter case (listed in appendix \ref{app:reference-configs}), as are the observational uncertainty and cadence. The orbit is adjusted for the Jupiter case by setting a perijove distance of $r_p=5$ Jupiter radii (compared to perigee radius $r_p=5$ Earth radii for the Earth encounter). We perform two fits for two different values of $v_\infty$: in one case we keep the Earth-encounter value of $v_\infty = 6$ km s$^{-1}$, and in the other case we increase the velocity such that the eccentricity (and therefore the angle between the orbits' hyperbolic asymptotes) are the same for both encounters. The latter requires $v_\infty = \sqrt{G\mu_M/r_p (e-1)}=32$ km s$^{-1}$. The rationale for these excess velocity values is that, for $v_\infty = 6$ km s$^{-1}$, we keep a physically typical encounter excess velocity at the expense of changing the orbit shape, whereas for $v_\infty =32$ km s$^{-1}$ the orbit shape is unchagned, although the asteroid speed is dramatically increased.

\jtd{Write}

\begin{table}
  \centering
  \begin{tabular}{c|cc}
    \hline 
    $K_{\ell m}$ & $v_\infty=6$ km $s^{-1}$ & $v_\infty = 32$ km s$^{-1}$\\ \hline 
    $\gamma_0$ & & 1.58 \\
    $K_{22}$ & & 2.32 \\
    $K_{20}$ & & 11.2 \\
    $\Re K_{33}$ & & 17.5 \\
    $\Im K_{33}$ & & 18.0 \\
    $\Re K_{32}$ & & 17.7 \\
    $\Im K_{32}$ & & 18.3 \\
    $\Re K_{31}$ & & 24.9 \\
    $\Im K_{31}$ & & 10.4 \\
    $K_{30}$ & & 51.8 \\ \hline
  \end{tabular}
  \caption{Ratio of posterior uncertainty for all density moments $K_{\ell m}$ between an Earth encounter and a Jupiter encounter with identical properties except for an increased perigee and an excess velocity $v_\infty$ given as the column headers.}
  \label{tab:jupiter-uncertainty}
\end{table}

There are additional effects of central body mass which are not captured in this analysis. For example, encounters with massive planets are more plentiful, so that observation for a fixed period of time will lead to a larger number of observed encounters conducive to low-uncertainty moment extraction (large $a_m$, small $r_p$, etc.). The distribution of $r_p$ in this encounter sample will also change; the ratio of the orbit impact parameter (the distance between the orbit asymptotes and the central body) to perigee distance is 
\begin{equation}
  \frac{b}{r_p} = \sqrt{1+2\frac{G\mu_M}{r_p v_\infty^2}}.
\end{equation}
It was mentioned above that the observable perigee distance $r_p$ increases with $a_M \sim \mu_M^{1/3}$, so that $\frac{b}{r_p}$ is larger for more massive planets and therefore the same impact parameter leads to lower periapse. Other effects, such as a change in the physical properties of the encountering asteroids, may also affect the fit uncertainties. However, the two effects we mention here both increase the number of encounters with properties conducive to accurate density moment extraction.

>>>>

We assess the feasibility of extracting density moments and density distributions for a general asteroid encounters, and ask what encounter properties are necessary to obtain the most information about asteroid density.

We derived a novel, arbitrary-order equation for the tidal torque experienced by an asteroid during an encounter with a planet of arbitrary shape and mass distribution. The tidal torque (equation \ref{eqn:tidal-torque}) is a function only of the density moments of the asteroid $K_{\ell m}$, the asteroid orbit, and the initial asteroid orientation. This equation revealed that the rotational velocity of the asteroid over time depends strongly on $K_{\ell m}$ and the initial orientation.

We built a fast simulation for an asteroid flyby, truncating the equation at second order, and used the simulation to extract first- and second-order density moments from synthetic spin pole data via a Markov Chain Monte Carlo fit. Other studies have extracted the first-order moments from encounters, but not second order. We observe the following general properties of the moments:
\begin{enumerate}
  \item The second-order moments $K_{3m}$ are generally a factor of $~\frac{a_m}{r_p}$ more uncertain than the first-order moments, where $a_m$ is the asteroid radius and $r_p$ is the perigee distance. For our reference asteroid, this fraction was about $10^{-5}$.
  \item Of the seven second-order moments, those that measure small-scale variations in density (i.e., $|m|$ is large) are generally less uncertain.
  \item The oblateness of the central body cannot be ignored without generating incorrect density moment estimates, but a nonzero oblateness does not greatly increase the precision of the extracted moments except perhaps when a moon is present.
\end{enumerate}

We assessed the posterior uncertainty generated by the fit process by adjusting eleven parameters of an asteroid on an Earth flyby. The main conclusions are listed below.

\begin{enumerate}
  \item Highly precise rotational velocity data is required to produce precise  moments. Especially the uncertainty on the asteroid rotational period must be small compared to the uncertainty on asteroid spin pole. Our reference asteroid, with $\sigma P/P = 10^{-5}$ and the spin pole known to within a few degrees of error, was accurate enough.
  \item A close encounter is vital for accurate second-order moment extraction, though some of the second-order moments can still be required for more distant encounters, extending to about a third of a Lunar distance for the reference asteroid.
  \item The asteroid rotational period must not be too small (smaller than about 5 hours, although $K_{30}$ will become unresolved at for periods smaller than 8 hours). Nor must the asteroid's rotational velocity point in an inconvenient direction (orthogonal to the orbital plane, or in the orbital plane and perpendicular to the perigee vector).
  \item The cadence of observation should be shorted than about 30 minutes (except $K_{30}$ will become unresolved for cadence worse than 10 minutes). Gaps of data can appear, specifically at perigee, but only for a little over an hour given 2 minute cadence before the posterior uncertainty significantly increases.
\end{enumerate}

Finally, we present three fast, linear methods for extracting density distributions (together with uncertainties) from the density moments and the shape of the asteroid as obtained by light-curve analysis. The extracted density distributions largely are consistent with the true distributions in all cases we tested. Slight errors in $K_{3m}$ dramatically change the overall distribution, so that they are not very precise. Indeed, the models sometimes predict negative density (though more complicated models could be made not to). Errors in the shape estimate of teh asteroid also distort the density distribution. Nevertheless, large-scale properties of the asteroid's true distribution, such as whether the mass is located in the middle or on the edges, are often observable. Two of the models presented (the harmonic and lumpy models) are flexible in their number of parameters, so that if only these large-scale properties are necessary, a small set of parameters can be chosen to prevent over-fitting.

In the appendices, we answer more specific questions about the behavior of the tidal torque system and the fit process that might arise.

@@

The physical parameters of the asteroid body are kept the same as the Earth encounter case (listed in appendix \ref{app:reference-configs}), as are the observational uncertainty and cadence. The orbit is adjusted for the Jupiter case by setting a perijove distance of $r_p=5$ Jupiter radii (compared to perigee radius $r_p=5$ Earth radii for the Earth encounter). We perform two fits for two different values of $v_\infty$: in one case we keep the Earth-encounter value of $v_\infty = 6$ km s$^{-1}$, and in the other case we increase the velocity such that the eccentricity (and therefore the angle between the orbits' hyperbolic asymptotes) are the same for both encounters. The latter requires $v_\infty = \sqrt{G\mu_M/r_p (e-1)}=32$ km s$^{-1}$. The rationale for these excess velocity values is that, for $v_\infty = 6$ km s$^{-1}$, we keep a physically typical encounter excess velocity at the expense of changing the orbit shape, whereas for $v_\infty =32$ km s$^{-1}$ the orbit shape is unchanged, although the asteroid speed is dramatically increased.

Section \ref{sec:scan-orbit} found that $v_\infty$ does not affect the posterior uncertainties, and indeed we find very little difference between the two $v_\infty$ cases. The ratio between the posterior uncertainties in the Jupiter and the Earth flybys are shown in table \ref{tab:jupiter-uncertainty}. In all cases, the Jupiter posteriors are more uncertain than Earth posteriors.

\begin{table}
  \centering
  \begin{tabular}{c|cc}
    \hline 
    $K_{\ell m}$ & $\sigma_\text{Jupiter}/\sigma_\text{Earth}$\\ \hline 
    $\gamma_0$ & 1.6 \\
    $K_{22}$ & 2.3 \\
    $K_{20}$ & 11 \\
    $\Re K_{33}$ & 18 \\
    $\Im K_{33}$ & 18 \\
    $\Re K_{32}$ & 18 \\
    $\Im K_{32}$ & 18 \\
    $\Re K_{31}$ & 25 \\
    $\Im K_{31}$ & 10 \\
    $K_{30}$ & 53 \\ \hline
  \end{tabular}
  \caption{Ratio of posterior uncertainty for all density moments $K_{\ell m}$ between an Earth encounter and a Jupiter encounter with identical properties except for an increased perigee.}
  \label{tab:jupiter-uncertainty}
\end{table}

This uncertainty can be understood as follows. The leading order of tidal torque is proportional to $\mu_m / D^3$. If $D/a_M$ (the ratio of the encounter distance to the central body radius) is roughly constant (as in this case, where $r_p/a_M=5$), then $\mu_m / D^3 \propto \rho_M$ where $\rho_M$ is the density of the central body. Therefore, little advantage is to be gained by looking for encounters of a massive planet in this sense. Since Jupiter is less dense than Earth, we expect that uncertainty in the first order parameters $\gamma_0$ and $K_{2m}$ would be slightly worse than in the case of Earth, which is seen in table \ref{tab-jupiter-uncertainty}.

Additionally, the second-order terms are damped by an additional factor of $a_m/D$, which decreases if a massive central body is used. Since Jupiter is about 10 times larger in radius than Earth, we expect that the $K_{3m}$ terms are about ten times more uncertain than the $K_{2m}$ components, which is the case.

The $K_{\ell 0}$ components violate the norm, with $\sigma_\text{Jupiter}/\sigma_\text{Earth}$ about five times greater for $K_{\ell 0}$ than other moments of the same $\ell$. In fact, $K_{30}$ essentially fills the prior. The special properties of $K_{\ell 0}$ have been analyzed noted before, in that they are always more uncertain than other moments and only enter certain parts of the tidal torque equation. \jtd{I did note this, right?} Table \ref{tab:jupiter-uncertainty} shows that these effects strengthen when a more massive central body is used.

There are additional effects of central body mass which are not captured in this analysis. For example, encounters with massive planets are more plentiful, so that observation for a fixed period of time will lead to a larger number of observed encounters conducive to low-uncertainty moment extraction (large $a_m$, small $r_p$, etc.). The distribution of $r_p$ in this encounter sample will also change; the ratio of the orbit impact parameter (the distance between the orbit asymptotes and the central body) to perigee distance is 
\begin{equation}
  \frac{b}{r_p} = \sqrt{1+2\frac{G\mu_M}{r_p v_\infty^2}}.
\end{equation}
It was mentioned above that the observable perigee distance $r_p$ increases with $a_M \sim \mu_M^{1/3}$, so that $\frac{b}{r_p}$ is larger for more massive planets and therefore the same impact parameter leads to lower periapse. Other effects, such as a change in the physical properties of the encountering asteroids, may also affect the fit uncertainties.

Table \ref{tab:jupiter-uncertainty} shows that an encounter with Jupiter yields less precise posteriors than an encounter with Earth, all else being equal. However, we also mention biases that will cause more large-$a_m$ and small-$r_p$ asteroids to encounter Jupiter, and these will yield increased precision. Which of these contradicting effects dominates depends on the specific asteroid population near Jupiter.
